---
title: "Format sorting phase and prepare models for HMSC"
output:
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: svg
    keep_html: yes
    html_preview: yes
---

Load and format species, experiment design, trait and phylogeny data for use with HMSC

# LIBRARIES
```{r eval=TRUE, include=FALSE}
library(here)
library(tidyverse)
library(ape)
library(Hmsc)

set.seed(341)
```

# PREPARE ABUNDANCE AND COVARIATE DATA
```{r}
counts <- read_tsv(here::here("data", "norm_counts.tsv")) %>%
  mutate(cil=factor(cil), wrm=factor(wrm)) %>%
  group_by(sample) %>%
  mutate(effort=log(sum(count.correct))) %>%
  ungroup() %>%
  mutate(Y=log(count.correct+1)) %>%
  mutate(Y=ifelse(Y==0, NA_real_, Y))
```

# FORMAT Y, X, SD MATRICES
Includes days 5, 9, 13. This is the sorting phase identified by the change point analysis
```{r}
#These species are too rare to model accurately
rm_sort <- c("HAMBI-0262")

counts_sort <- counts %>%
  filter(day < 17 & day != 0) %>%
  filter(!(strainID %in% rm_sort)) %>%
  mutate(strainID=factor(strainID)) 
```

# Y MATRIX 
These are species abundances

## ABUNDANCE CONDITIONAL ON PRESENCE
Ya is an abundance matrix but we set 0 values == NA
```{r}
Ya <- counts_sort %>%
  select(sample, strainID, Y=Y) %>%
  pivot_wider(names_from="strainID", values_from="Y") %>%
  column_to_rownames(var="sample") %>%
  as.matrix()
```

## PRESENCE/ABSENCE
Yp in the richness matrix of just 0/1
```{r}
Yp <- counts_sort %>%
  select(sample, strainID, Y=PA) %>%
  pivot_wider(names_from="strainID", values_from="Y") %>%
  column_to_rownames(var="sample") %>%
  as.matrix()
```

# X MATRIX 
```{r}
XData <- counts_sort %>%
  select(sample, day, cil, wrm, effort) %>% #replicate, treatment
  distinct() %>%
  column_to_rownames(var="sample") %>%
  as.data.frame()
```

# STUDY DESIGN
```{r}
studyDesign <- counts_sort %>% 
  select(sample, day.coord=day, microcosmID.coord=microcosmID) %>%
  distinct() %>%
  mutate(microcosmID.unit=factor(paste0("microcosmID_", microcosmID.coord)),
         day.unit=factor(paste0("day_",day.coord))) %>%
  column_to_rownames(var='sample') %>%
  as.data.frame()
```

# TRAIT DATA
```{r}
scale2 <- function(x, na.rm = FALSE) (x - mean(x, na.rm = na.rm)) / sd(x, na.rm)

TrData <- read_tsv(here::here("data", "HAMBI-traits.tsv")) %>%
  select(strain, D, n_carbon_ecoplate, r_specific_growth_rate, biofilm_formation) %>%
  column_to_rownames(var="strain") %>% 
  mutate(across(everything(), scale2)) %>% 
  as.data.frame()

TrData <- TrData[!(row.names(TrData) %in% rm_sort), ]
```

# PHYLOGENETIC TREE
```{r}
phyloTree <- read.tree(here::here("data", "phylotree.nwk"))
phyloTree <- drop.tip(phyloTree, rm_sort)
```

# FORMAT FOR RANDOM EFFECTS
```{r}
time <- as.matrix(data.frame(day=unique(studyDesign$day.coord)))
rownames(time) <- unique(studyDesign$day.unit)
rLday <- HmscRandomLevel(sData = time)

microcosmID <- as.matrix(data.frame(microcosmID=unique(studyDesign$microcosmID.coord)))
rownames(microcosmID) <- unique(studyDesign$microcosmID.unit)
rLmicrocosmID <- HmscRandomLevel(units = microcosmID)

studyDesign$day <- studyDesign$day.unit
studyDesign$microcosmID <- studyDesign$microcosmID.unit
studyDesign <- studyDesign[c(5,6)]
```

# SPECIFY MODEL
Here we will fit three different types of models.

1. A "FULL" model that includes both predation regime (environmental), time, and sequencing effort as fixed covariates as well as temporal random effects and hierarchical/repeated measure random effects.

2. A "ENVI" model that includes predation regime (environmental), time, and sequencing effort as fixed covariates but does not account for random effects.

3. A "TIME" model that excludes environmental covariates (includes only time and sequencing effort) and only models baseline abundance, but includes temporal and hierarchical/repeated measure random effects.

In the sorting phase we also included time as a fixed effect. In the shannon diversity data we saw a linear drop in diversity in the first 17 days. This is also mirrored in the abundances of individual species where some species rapidly dropped in log abundance while other increased. Including time as a fixed effect allows us to capture this general temporal dynamic while not blowing up the model with many temporal terms like polynomials or splines.

## HURDLE MODELS
Ecological count data are often dominated by zeros, i.e. they are zero-inflated. One standard model that can be fitted to such data is the Zero-Inflated Poisson model. While HMSC-R 3.0 does not include this or other zero-inflated models, we note that it is always possible to apply the closely-related Hurdle approach. In the Hurdle model, one analyses separately occurrence data 1*(Y>0) and abundance conditional on presence Y[Y==0] = NA. We note that these two aspects of the data (occurrence and abundance conditional on presence) are statistically independent of each other, and thus it is interesting to compare results obtained for them, e.g. to see if variation in occurrence and variation in abundance are explained by the same environmental covariates. In HMSC-R 3.0, occurrence data are analysed by the probit model, whereas the best model for abundance (conditional on presence) depends on the type of the data.

# MAKE MODEL OBJECTS

## ABUNDANCE MODELS 
```{r}
ma_full <- Hmsc(Y           = Ya, 
                XData       = XData, 
                XFormula    = ~day*cil*wrm + effort, 
                TrData      = TrData,
                TrFormula   = formula(paste0("~",paste(names(TrData), collapse = "+"))),
                phyloTree   = phyloTree,
                studyDesign = studyDesign,
                ranLevels   = list(day=rLday, microcosmID=rLmicrocosmID),
                distr       = "normal",
                YScale      = TRUE)

ma_envi <- Hmsc(Y           = Ya, 
                XData       = XData, 
                XFormula    = ~day*cil*wrm + effort, 
                TrData      = TrData,
                TrFormula   = formula(paste0("~",paste(names(TrData), collapse = "+"))),
                phyloTree   = phyloTree,
                studyDesign = studyDesign,
                distr       = "normal",
                YScale      = TRUE)

ma_time <- Hmsc(Y           = Ya, 
                XData       = XData, 
                XFormula    = ~day + effort, 
                TrData      = TrData,
                TrFormula   = formula(paste0("~",paste(names(TrData), collapse = "+"))),
                phyloTree   = phyloTree,
                studyDesign = studyDesign,
                distr       = "normal",
                YScale      = TRUE)

saveRDS(ma_full, here::here("models_unfit", "sort_ma_full.rds"))
saveRDS(ma_envi, here::here("models_unfit", "sort_ma_envi.rds"))
saveRDS(ma_time, here::here("models_unfit", "sort_ma_time.rds"))
```

## PROBIT MODELS 
```{r}
mp_full <- Hmsc(Y           = Yp, 
                XData       = XData, 
                XFormula    = ~day*cil*wrm + effort, 
                TrData      = TrData,
                TrFormula   = formula(paste0("~",paste(names(TrData), collapse = "+"))),
                phyloTree   = phyloTree,
                studyDesign = studyDesign,
                ranLevels   = list(day=rLday, microcosmID=rLmicrocosmID),
                distr       = "probit",
                YScale      = TRUE)

mp_envi <- Hmsc(Y           = Yp, 
                XData       = XData, 
                XFormula    = ~day*cil*wrm + effort, 
                TrData      = TrData,
                TrFormula   = formula(paste0("~",paste(names(TrData), collapse = "+"))),
                phyloTree   = phyloTree,
                studyDesign = studyDesign,
                distr       = "probit",
                YScale      = TRUE)

mp_time <- Hmsc(Y           = Yp, 
                XData       = XData, 
                XFormula    = ~day + effort, 
                TrData      = TrData,
                TrFormula   = formula(paste0("~",paste(names(TrData), collapse = "+"))),
                phyloTree   = phyloTree,
                studyDesign = studyDesign,
                distr       = "probit",
                YScale      = TRUE)

saveRDS(mp_full, here::here("models_unfit", "sort_mp_full.rds"))
saveRDS(mp_envi, here::here("models_unfit", "sort_mp_envi.rds"))
saveRDS(mp_time, here::here("models_unfit", "sort_mp_time.rds"))
```

# TEST MODELS ARE CORRECT FORMAT
```{r eval=FALSE, include=TRUE}
m1 <- sampleMcmc(ma_envi, thin = 1, samples = 100, 
                     transient = 50, nChains = 2)
```
