---
title: "Multiple change point regression"
output:
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: svg
    keep_html: yes
    html_preview: yes
---

```{r}
library(tidyverse)
library(here)
library(mcp)
library(rjags)
```

# LOAD DATA
```{r}
predator <- read_tsv(here("data", "predator_density.tsv"), col_types="ccccdddd") %>%
  mutate(replicate=factor(replicate, levels=c("A", "B", "C", "D"))) %>%
  mutate(treatment=factor(treatment, levels=c("H", "HN", "HPanc", "HPevo", "HNPanc", "HNPevo")),
         microcosmID=factor(microcosmID))

counts <- read_tsv(here::here("data", "norm_counts.tsv"))
```

## LOAD DIVNET ESTIMATE
```{r}
div.sum <- readRDS(here::here("data", "diversity_summary.rds"))
```


# MULTIPLE CHANGE POINT REGRESSION
https://lindeloev.github.io/mcp/index.html

The idea here is that we assume each microcosm takes some amount of time to reach equilibrium dynamics. We want to do further analysis (HMSC) on each of these phases (pre equilirbium vs equilibrium) separately. We use the stabilization of species diversity as an indicator of equilibrium so we fit a two-segment change point regression model. The first segment models community diversity changing with time (the pre equilibrium phase). The second segment models diversity only as an intercept so that diversity can oscillates around a fixed mean value but that the mean doesn't change with time.

population mean for changepoint of 12.2438102 +- 3.1680302	days


```{r}
dpsegin <- div.sum %>% 
  filter(inference=="plugin")
```

## DEFINE MODELS

### MODEL 00
segment 1: change in intercept and slope modeled on day

segment 2: population-level change point with treatment-varying change points sampled around the population mean.joined segment at cp_1. Slope = 0. Intercept (Int) is estimated

```{r}
model00 <- list(
 estimate ~ 1 + day, 
 estimate ~ 1 + (1|treatment) ~ 0  
)

fit00 <- mcp(model00, data = dpsegin, sample = "both", iter = 10000)

fit00$loo <- loo(fit00)
```

Summary of the model fit
```{r}
summary(fit00)
```

Summary of random effects. Treatment-specific deviation from the population mean change point
```{r}
ranef(fit00)
```
Plot
```{r}
mcpplot00 <- plot(fit00, facet_by = "treatment", q_predict=F)
mcpplot00
```

Check convergence
```{r}
plot_pars(fit00, pars = "varying", type = "trace", ncol = 3)
```

```{r}
plot_pars(fit00, regex_pars = "^cp_1_treatment", type = "dens_overlay", ncol = 3)
```

### MODEL 01
This is an alternative model where we allow for a change in slope in segment 2.

segment 1: change in intercept and slope modeled on day (same as before)

segment 2: population-level change point with treatment-varying change points sampled around the population mean. Joined segment at cp_1. Slope modeled on day. Intercept (Int) is estimated

```{r}
model01 = list(
 estimate ~ 1 + day,
 estimate ~ 1 + (1|treatment) ~ 0 + day
)

fit01 <- mcp(model01, data = dpsegin)

fit01$loo = loo(fit01)
```

```{r}
summary(fit01)
```

```{r}
ranef(fit01)
```

```{r}
mcpplot01 <- plot(fit01, facet_by = "treatment", q_predict=F)
mcpplot01
```

### MODEL 02
This is an alternative model where we allow for a population changepoint only. No-treatment specific effects

segment 1: change in intercept and slope modeled on day

segment 2: population-level change point only. Joined segment at cp_1. Slope modeled on day. Intercept (Int) is estimated

```{r}
model02 = list(
 estimate ~ 1 + day,
 estimate ~ 1 ~ 0 + day
)

fit02 <- mcp(model02, data = dpsegin)

fit02$loo = loo(fit02)
```

```{r}
summary(fit02)
```

```{r}
mcpplot02 <- plot(fit02, q_predict=F)
mcpplot02
```

## COMPARE MODELS
```{r}
loo::loo_compare(fit00$loo, fit01$loo, fit02$loo)
```

We can use the cross-validation from the loo package to compare the predictive performance of mcp models. Use loo to compute Widely Applicable Information Criterion (WAIC) or Estimated Log Predictive Density (ELPD) for each model, and then compare them using loo::loo_compare().

So the first model (“model00”) is prefered since model01 has a smaller ELPD. We conclude that a non-zero slope in the second "quasi-equilibrium" phase does not improve out-of-sample predictions over a zero slope (that is constant longitudinal mean diversity).

## HYPOTHESIS TEST
```{r}
hypothesis(fit00, c("`cp_1_treatment[H]` = 0",
                    "`cp_1_treatment[HN]` = 0",
                    "`cp_1_treatment[HPanc]` = 0",
                    "`cp_1_treatment[HPevo]` = 0",
                    "`cp_1_treatment[HNPanc]` = 0",
                    "`cp_1_treatment[HNPevo]` = 0"
                    ))
```

```{r}
mcpplotfinal <- plot(fit00, facet_by="treatment", q_predict=TRUE)

thresh <- tibble::tribble(
  ~treatment, ~day,       ~event,
        "HN",  17, "stabilized",
     "HPanc",  20, "stabilized",
     "HPevo",  20, "stabilized",
    "HNPanc",  21, "stabilized",
    "HNPevo",  25, "stabilized",
    "HNPevo",  32,       "lost",
    "HNPanc",  11,       "lost"
  ) %>%
  mutate(event=factor(event))

p5 <- mcpplotfinal +
  facet_grid(treatment ~ .) +
  geom_pointrange(data=dpsegin, aes(x=day, y=estimate, ymin = lower, ymax = upper), fatten=1) + 
  geom_vline(data=thresh, aes(xintercept=day, linetype=event)) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

p5
```
