---
title: "Fig. 3"
output:
  github_document:
    toc: yes
    toc_depth: 2
    fig_width: 7
    fig_height: 5
    dev: svg
    keep_html: yes
    html_preview: yes
---

```{r}
library(tidyverse)
library(here)
library(rcartocolor)
library(patchwork)
library(scales)
library(ggforce)
library(vegan)
library(withr)
library(cvequality)

`%nin%` = Negate(`%in%`)
```

# LOAD DATA
```{r message=FALSE}
predator <- read_tsv(here("data", "predator_density.tsv"), col_types="ccccdddd") %>%
  mutate(replicate=factor(replicate, levels=c("A", "B", "C", "D"))) %>%
  mutate(treatment=factor(treatment, levels=c("H", "HN", "HPanc", "HPevo", "HNPanc", "HNPevo")),
         microcosmID=factor(microcosmID))

counts <- read_tsv(here::here("data", "norm_counts.tsv"))
```



# SPECIES ABUNDANCE DISTRIBUTION

## FOR ORDERING STRAINS BY ABUNDANCE
```{r}
strainrecode <- counts %>%
  filter(day > 0) %>%
  group_by(treatment, day, replicate, strainID) %>%
  summarize(spsum=sum(count.correct)) %>%
  ungroup() %>%
  group_by(treatment, day, replicate) %>%
  mutate(nreads=sum(spsum)) %>%
  ungroup() %>%
  mutate(RA=round(spsum/nreads, 2)) %>%
  group_by(strainID) %>%
  summarize(mRA=mean(RA), sdRA=sd(RA)) %>%
  ungroup() %>%
  arrange(mRA) %>%
  distinct(strainID) %>%
  rowid_to_column("strain.num")
  
strainrecode <- counts %>% 
  group_by(strainID) %>%
  summarize(spsum=sum(count.correct)) %>%
  ungroup() %>%
  mutate(RA=spsum/sum(spsum)) %>%
  arrange(desc(RA)) %>%
  distinct(strainID) %>%
  rowid_to_column("strain.num")
```

Setup tibble
```{r}
sad <- counts %>%
  mutate(count.correct=ifelse(count.correct==0, 1, count.correct)) %>%
  filter(day > 0) %>%
  group_by(treatment, day, replicate, strainID) %>%
  summarize(spsum=sum(count.correct)) %>%
  ungroup() %>%
  group_by(treatment, day, replicate) %>%
  mutate(nreads=sum(spsum)) %>%
  ungroup() %>%
  mutate(RA=round(spsum/nreads, 2)) %>%
  group_by(treatment, strainID) %>%
  summarize(mRA=mean(RA), sdRA=sd(RA)) %>%
  mutate(mRA=ifelse(mRA < 1e-4, 1e-4, mRA)) %>%
  ungroup() %>%
  mutate(ymin=mRA-sdRA, ymax=mRA+sdRA) %>%
  mutate(ymin=ifelse(ymin<=0, 1e-4, ymin)) %>%
  left_join(., strainrecode) %>%
  mutate(treatment=factor(treatment, levels=c("H", "HPanc", "HPevo", "HN", "HNPanc", "HNPevo")))
```

plot
```{r}
p0 <- ggplot(sad) + 
  geom_linerange(aes(x=strain.num, ymin = ymin, ymax = ymax), color="grey80") + 
  geom_point(aes(x=strain.num, y=mRA), size=0.5) + 
  facet_grid(~ treatment) +
  labs(y="", x="") +
  #coord_flip() + 
  #scale_y_continuous(trans = trans_reverser('log'), breaks=c(0.5, 0.1, 0.01)) +
  scale_y_continuous(trans = 'log10',
                       breaks = c(1, 0.5, 0.1, 0.01, 0.001),
                       labels = label_percent(accuracy = 0.1)) + 
  # scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1e-4, base = 10),
  #                      breaks = c(1, 0.5, 0.1, 0.01, 0.001),
  #                      labels = label_percent(accuracy = 0.1)) + 
  scale_x_continuous(n.breaks=24, limits = c(0,25), expand=c(0,0)) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

p0
```

# HEAT MAPS
```{r}
my_colors <- carto_pal(6, "Vivid")

names(my_colors) <- c("HPanc", "HPevo", "HN", "HNPanc", "HNPevo", "H")

my_colors
```

Calculate means over replicates
```{r}
hmapdf.mn <- counts %>% 
  mutate(count=exp(rlogcount)) %>%
  dplyr::select(-lograrecount, -log10rab, -sflogcount, -rlogcount, -PA) %>%
  arrange(strainID, microcosmID, day) %>%
  group_by(treatment, strainID, day) %>%
  mutate(count.mean=mean(count)) %>% ungroup() %>%
  dplyr::select(day, treatment, cil, wrm, strainID, count.mean) %>%
  distinct() %>%
  group_by(treatment, strainID) %>%
  mutate(count.mean.f=count.mean/max(count.mean)) %>%
  mutate(count.mean.f1=log(count.mean/max(count.mean))/sqrt(2)) %>%
  mutate(xmin=day, xmax=lead(day)) %>%
  mutate(xmax=ifelse(is.na(xmax), 65, xmax)) %>%
  ungroup() %>%
  left_join(., strainrecode) %>%
  mutate(treatment=factor(treatment, levels=c("H", "HPanc", "HPevo", "HN", "HNPanc", "HNPevo")))
```

```{r}
p1 <- ggplot(hmapdf.mn) +
  geom_rect(aes(ymin=xmin, ymax=xmax, xmin = strain.num-0.5, 
                xmax = strain.num+0.5, fill = count.mean.f)) +
  geom_hline(yintercept=32) +
  geom_hline(yintercept=11) +
  scale_fill_viridis_c(option = "C", begin = 0.05, end=0.95, direction =-1,
                       trans = scales::pseudo_log_trans(sigma = 1e-3, base = exp(1)),
                       breaks = breaks_log(n = 5, base = 10),
                       labels = label_percent(accuracy = 0.1)) +
  facet_grid(~ treatment) +
  labs(y="", x="") +
  scale_x_continuous(n.breaks=24, limits = c(0,25), expand=c(0,0)) + 
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

p1
```

# ORDINATION
```{r}
a <- counts %>%
  filter(day != 0) %>%
  select(sample, strainID, count.correct) %>%
  group_by(sample) %>%
  mutate(count.correct=count.correct/sum(count.correct)) %>%
  ungroup() %>%
  pivot_wider(names_from="strainID", values_from="count.correct") %>%
  column_to_rownames(var="sample") %>%
  data.frame()

# Bray-curtis dissimilarity 
with_preserve_seed(dist <- vegdist(a,  method = "bray"))      

# do ordination
with_preserve_seed(nmds_ord <- metaMDS(dist, k = 2, trymax = 100, trace = F, autotransform = FALSE))

# tibble of all samples
nmds2plot <- data.frame(nmds_ord$points) %>%
  rownames_to_column(var="sample") %>%
  left_join(., distinct(select(counts, sample:wrm))) %>%
  mutate(method="NMDS") %>%
  dplyr::rename(axis_1 = MDS1, axis_2 = MDS2) %>%
  mutate(treatment=factor(treatment, levels=c("H", "HPanc", "HPevo", "HN", "HNPanc", "HNPevo")))

# get centroids for each treatment/day
centroids <- nmds2plot %>% 
  group_by(treatment, day) %>%
  mutate(NMDS1=mean(axis_1), NMDS2=mean(axis_2),
                NMDS1sd=sd(axis_1), NMDS2sd=sd(axis_2)) %>%
  ungroup() %>%
  mutate(treatment=factor(treatment, levels=c("H", "HPanc", "HPevo", "HN", "HNPanc", "HNPevo")))

# for making spider plot
spiders <- nmds2plot %>% 
  select(sample, fromx=axis_1, fromy=axis_2) %>%
  left_join(., centroids) %>%
  dplyr::rename(tox=NMDS1, toy=NMDS2) %>%
  mutate(treatment=factor(treatment, levels=c("H", "HPanc", "HPevo", "HN", "HNPanc", "HNPevo")))

centroids <- centroids %>%
  select(NMDS1, NMDS2, NMDS1sd, NMDS2sd, day, treatment) %>% distinct()

p2 <- ggplot() + 
  geom_point(data=nmds2plot, aes(x=axis_1, y=axis_2, color=factor(day)), size=0.5) +
  geom_segment(data=spiders, aes(x=fromx, xend = tox, y=fromy, yend = toy,
                                 group=treatment, color=factor(day))) +
  geom_link2(data = centroids, aes(x=NMDS1, y=NMDS2, color=factor(day), group=treatment)) +
  geom_point(data = centroids, aes(x=NMDS1, y=NMDS2, fill=factor(day)), size=1, shape=21) + #size=((NMDS1sd+NMDS2sd)/2)^-1
  facet_grid(~treatment) + 
  scale_color_brewer(type="qual", palette="Paired") +
  scale_fill_brewer(type="qual", palette="Paired") +
  theme_bw() +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.background = element_blank())

p2
```

# COMBINED PLOT
```{r}
pfinal <- (p0 / p1 / p2) + plot_layout(guides = 'collect', 
                                     widths = c(1,1,1),
                                     heights= c(2,4,3),
                                     ncol=1)

pfinal
```

```{r}
ggsave(here("figs", "fig3.svg"), pfinal, width=20.8, height=12, units="cm",
       device="svg")
```


# JUMP LENGTHS
see [here](https://www.sciencedirect.com/science/article/pii/S1369527418300092#bib0070)

Are the jumps through community space larger or smaller for a given treatment?

```{r}
# calculate distances by pythagorean theorem
jumps <- nmds2plot %>%
  arrange(treatment, replicate, day) %>%
  group_by(microcosmID) %>%
  mutate(axis1_lag = lag(axis_1),
         axis2_lag = lag(axis_2)) %>%
  mutate(d=sqrt((axis_1-axis1_lag)^2 + (axis_2-axis2_lag)^2)) %>%
  ungroup()
```

```{r}
jumps %>%
  group_by(treatment) %>%
  summarize(l=mean(d, na.rm=T)) %>%
  arrange(l)
```

```{r}
# what does it look like
ggplot(jumps, aes(x=treatment, y=d, color=factor(day))) + 
  geom_point()
```

```{r}
# choosing best model. Gamma is best
m0 <- glm(data=jumps, formula=d ~ treatment, family=gaussian())
m1 <- glm(data=jumps, formula=d ~ treatment, family=Gamma())
m2 <- glm(data=jumps, formula=d ~ treatment, family=inverse.gaussian())
```

```{r}
library(performance)
compare_performance(m0, m1, m2, rank = TRUE)
```

```{r}
check_model(m1)
```

```{r}
summary(m1)
```

```{r}
library(rstanarm)
library(bayestestR)
library(modelbased)

m2stan <- stan_glm(d ~ treatment,
                  data = jumps,
                  family=Gamma(),
                  iter = 4000,
                  cores = 4,
                  seed = 12345)
```


```{r}
describe_posterior(m2stan, test = c("pd", "rope"), ci = 0.95,
                   rope_range = c(-0.1, 0.1), rope_ci=1) %>%
  as_tibble() %>%
  mutate(Median=round(Median, 2),
         CI=paste0("[", round(CI_low, 2),", ",round(CI_high, 2),"]"),
         pd=pd*100,
         ROPE_Percentage=ROPE_Percentage*100) %>%
  select(Parameter, Median, CI, pd, ROPE_low, ROPE_Percentage) #%>%
  xtable::xtable()
```

```{r}
estimate_means(m2stan, transform = "none")
```

Contrasts
```{r}
estimate_contrasts(m2stan, 
                levels="treatment", 
                test=c("p_direction", "rope"), 
                ci = 0.95,
                rope_range = c(-0.1, 0.1),
                rope_ci = 1)# %>%
  tibble() %>%
  filter(pd > 0.97) %>%
  mutate(Difference=round(Difference, 2),
         CI=paste0("[", round(CI_low, 2),", ",round(CI_high, 2),"]"),
         pd=pd*100) %>%
  mutate(ROPE_low=0.1,
         ROPE_Percentage=ROPE_Percentage*100,
         treatment=paste(Level1, Level2, sep=" - ")) %>%
  select(treatment, Difference, CI, pd, ROPE_low, ROPE_Percentage) %>%
  xtable::xtable()
```


Default community with no predator has greatest jump lengths

So CHTV, N, and HNCHTV all have smaller jumps through community space than HPanc

Are the treatments differentially variable? Does one condition vary more than we expect by chance
```{r}
jumps1 <- jumps %>% select(d, treatment) %>% drop_na() %>% data.frame()
with(jumps1, asymptotic_test(d, treatment))
with(jumps1, mslr_test(nr = 1e4, d, treatment))
```
No evidence that they are differentially variable